A Mechanised Library for Generic (Co)Programming

### 2024-03-06 paper/paper.tex
Recursive anamorphisms are hylomorphisms on the initial algebra.

The seed function can be modelled by a coalgebra. For example, for a type of
states \coq{X} and an alphabet type \coq{L} we can define a (labelled)
transition system (LTS) on \coq{X} as a function \coq{c : X -> L * X}
implementing the \emph{transition} map. In particular, for a
state \coq{x1 : X}, \coq{c x1} returns a pair \coq{(l, x2) : L * X}
where \coq{l : L} is the observable action and \coq{x2 : X} is the next state.
Note that for a generic functor \coq{F} an \coq{F}-coalgebra on the
carrier \coq{X} is a map \coq{X -> FX}. In the previous example \coq{c} is
an \coq{F}-coalgebra on the carrier \coq{X} with \coq{FX = L * X}.

@@

Thus, in general given an $F$-algebra and $F$-coalgebra, the hylomorphism is the unique solution to the equation
\begin{equation}
  \label{eq:hylo}
  f = a \circ F\; f \circ c
\end{equation}
However this solution does not necessarily exists for an arbitrary pair of
algebras and coalgebras.  In fact, this definition cannot be accepted by Coq,
since there is no structurally smaller argument to do recursion on and there is
no corecursive definition that is guaranteed to be syntactically guarded -- as
required by Coq -- due to the use of an arbitrary algebra $a$. Furthermore,
solutions to the hylomorphism equation are \emph{not} guaranteed to be unique,
or to even exist. To address these issues, we mechanise \textbf{recursive
hylomorphisms} which are guaranteed to have a unique solution to the
hylomorphism equation.

If $c$ is a recursive coalgebra, then the equation (\ref{eq:hylo}) has a unique
solution. We show this by first providing the definition for recursive
hylomorphisms in Coq, and then proving that any function satisfying this
equation is a hylomorphism.
\begin{coqcode}
Definition hylo_def (a : Alg F B) (c : Coalg F A)
  : forall (x : A), RecF c x -> B := fix f x H
  := match c x as h0 return
       (forall e : Pos (shape h0), RecF c (cont h0 e)) -> B
     with
     | MkApp s_x c_x => fun H => a (MkApp s_x (fun e => f (c_x e) (H e)))
     end (RecF_inv H).
\end{coqcode}
The definition of recursive hylomorphisms is recursive on the proof that the
anamorphism can only be applied finitely many times on any input. We again use
\coq{RecF_inv} to obtain the structurally smaller proof to use in
the recursive calls.  As we did with catamorphisms and recursive anamorphisms,
we prove that \coq{hylo_def} is respectful, and use this proof to
built the corresponding higher-order proper morphism:
\begin{coqcode}
Definition hylo : Alg F B ~> RCoalg F A ~> A ~> B := ...
\end{coqcode}
Finally, we show that recursive hylomorphisms are the unique solution to
the hylomorphism equation.
\begin{coqcode}
Lemma hylo_univ (g : Alg F B) (h : RCoalg F A) (f : A ~> B)
    : f =e hylo g h <-> f =e g \o fmap f \o h.
\end{coqcode}
This result can be used to prove hylomorphism laws in a way that is analogous
to pen-and-paper program calculation techniques. We show the rule for
hylomorphism fusion as an example:
\begin{coqcode}
Lemma hylo_fusion_l (h1 : RCoalg F A) (g1 : Alg F B) (g2 : Alg F C)
  (f2 : B ~> C) (E2 : f2 \o g1 =e g2 \o fmap f2) : f2 \o hylo g1 h1 =e hylo g2 h1.

  rewrite hylo_univ, fmap_comp, compA, <- E2,
          <- compA, <- compA, (compA g1), <- hylo_unr.  reflexivity.
Qed.
  (* f2 \o hylo g1 h1 =e hylo g2 h1 *)
  (* f2 \o hylo g1 h1 =e (g2 \o fmap (f2 \o hylo g1 h1)) \o h1 *)
  (* f2 \o hylo g1 h1 =e (g2 \o (fmap f2 \o fmap (hylo g1 h1))) \o h1 *)
  (* f2 \o hylo g1 h1 =e ((g2 \o fmap f2) \o fmap (hylo g1 h1)) \o h1 *)
  (* f2 \o hylo g1 h1 =e ((f2 \o g1) \o fmap (hylo g1 h1)) \o h1 *)
  (* f2 \o hylo g1 h1 =e (f2 \o g1) \o (fmap (hylo g1 h1) \o h1) *)
  (* f2 \o hylo g1 h1 =e f2 \o (g1 \o (fmap (hylo g1 h1) \o h1)) *)
  (* f2 \o hylo g1 h1 =e f2 \o ((g1 \o (fmap (hylo g1 h1)) \o h1) *)
  (* f2 \o hylo g1 h1 =e f2 \o hylo g1 h1 *)
\end{coqcode}
Using the hylo fusion law, we can prove the well-known \emph{deforestation
optimisation}, where two consecutive recursive computations, one that builds a
data structure, and another one that consumes it, can be fused together into a
single recursive definition. This, in turn, allows us to prove that a recursive
hylomorphism is the composition of a catamorphism and a recursive anamorphism.
\begin{coqcode}
Lemma deforest (h1 : RCoalg F A) (g2 : Alg F C)
  (g1 : Alg F B) (h2 : RCoalg F B) (INV: h2 \o g1 =e id)
  : hylo g2 h2 \o hylo g1 h1 =e hylo g2 h1.

Corollary cata_ana_hylo `(F : Cont Sh P) `{setoid A} `{setoid B}
  (g : Alg F B) (h : RCoalg F A)
: cata g \o rana h =e hylo g h.
\end{coqcode}
This result will allow us to \emph{calculate} programs by providing small, easy
to verify implementations, and combining them into efficient recursive
functions by \emph{using Coq's proofs of hylomorphism laws as optimisation
passes}.

\mpav{Moved here}
In particular, we show that the terminal coalgebras
for $F$, coupled with a proof that they are finite are also initial
$F$-algebras:
\begin{coqcode}
Inductive FinF : GFix F -> Prop :=
| FinF_fold (x : GFix F) : (forall e, FinF (cont (g_out x) e)) -> FinF x.
\end{coqcode}
Catamorphisms from this initial algebra are denoted by \coq{ccata},
and they can be used to define the isomorphism between
\coq{{x : GFix F | FinF x}} and
\coq{LFix F}. In particular:
\begin{coqcode}
Lemma cata_ccata `{setoid A} (f : Alg F A) : cata f \o ccata l_in =e ccata f.
Lemma ccata_cata `{setoid A} (f : Alg F A) : ccata f \o cata lg_in =e cata f.
\end{coqcode}
Using this new initial algebra, we can compose catamorphisms and anamorphisms,
\emph{as long as the anamorphisms produce finite trees}. We refer the reader to
the supplementary material and Coq code to see how we encode this. In this
paper, however, we focus on the fact that, if anamorphisms produce finite trees,
then their return type can also be the \emph{least} fixed point of the container
extension. We encode in Coq the notion of \emph{recursive anamorphisms} and
\emph{recursive coalgebras}~\cite{AdamekMM19,CaprettaUV04}. Recursive
coalgebras are coalgebras whose anamorphism produces finite trees. The
following predicate captures this fact, \coq{h} is recursive if \coq{h} can only
be applied finitely many times:
\begin{coqcode}
Inductive RecF (h : Coalg F A) : A -> Prop :=
| RecF_fold x : (forall e, RecF h (cont (h x) e)) -> RecF h x.
\end{coqcode}
This predicate is equivalent to the proposition that the anamorphism of such
coalgebras produce only finite trees:
\begin{coqcode}
Lemma ana_rec_term `{setoid A} (h : Coalg F A)
  : (forall x, FinF (ana h x)) <-> (forall x, RecF h x).
\end{coqcode}
Recursive anamorphisms are implemented in Coq recursively on the structure of
the proof that they can only be applied finitely many times (\coq{RecF}).  Note
that such proofs will be erased when extracting code to OCaml, and this is
standard practice when building general recursive functions in
Coq\footnote{\dcas{This technique had a name in the community, but I cannot
remember. Something predicates}}. The core definitions are the following:
\begin{coqcode}
Lemma RecF_inv (h : Coalg F A) x : RecF h x -> forall e, RecF h (cont (h x) e).

>>>>

The seed function can be modelled by a coalgebra. For example, for a type of
states \coq{X} and an alphabet type \coq{L} we can define a (labelled)
transition system (LTS) on \coq{X} as a function \coq{c : X -> L * X}
implementing the \emph{transition} map. In particular, for a
state \coq{x1 : X}, \coq{c x1} returns a pair \coq{(l, x2) : L * X}
where \coq{l : L} is the observable action and \coq{x2 : X} is the next state.
Note that for a generic functor \coq{F} an \coq{F}-coalgebra on the
carrier \coq{X} is a map \coq{X -> FX}. In the previous example \coq{c} is
an \coq{F}-coalgebra on the carrier \coq{X} with \coq{FX = L * X}.
In our development we use the following notation for coalgebras:
\begin{coqcode}
Notation Coalg F A := (A ~> App F A) (only parsing).
\end{coqcode}

@@

Thus, in general given an $F$-algebra and $F$-coalgebra, the hylomorphism is the
unique solution to the equation
\begin{equation}
  \label{eq:hylo}
  f = a \circ F\; f \circ c
\end{equation}

However this solution does not necessarily exists for an arbitrary pair of
algebras and coalgebras. In fact, this definition cannot be accepted by Coq,
since there is no structurally smaller argument to do recursion on and there is
no corecursive definition that is guaranteed to be syntactically guarded -- as
required by Coq -- due to the use of an arbitrary algebra $a$.  Furthermore,
solutions to the hylomorphism equation are \emph{not} guaranteed to be unique,
or to even exist. To address these issues, we mechanise \emph{recursive
hylomorphisms} which are guaranteed to have a unique solution to the
hylomorphism equation. These are hylomoprhisms where the coalgebra is
\emph{recursive}.

Recursive coalgebras are coalgebras whose anamorphism produces finite data which
what the following predicate captures:
\begin{coqcode}
Inductive RecF (h : Coalg F A) : A -> Prop :=
| RecF_fold x : (forall e, RecF h (cont (h x) e)) -> RecF h x.
\end{coqcode}
Recursive anamorphisms are implemented in Coq recursively on the structure of
the proof for the type (\coq{RecF}) as follows:
\begin{coqcode}
Definition hylo_def (a : Alg F B) (c : Coalg F A)
  : forall (x : A), RecF c x -> B := fix f x H
  := match c x as h0 return
       (forall e : Pos (shape h0), RecF c (cont h0 e)) -> B
     with
     | MkApp s_x c_x => fun H => a (MkApp s_x (fun e => f (c_x e) (H e)))
     end (RecF_inv H).
\end{coqcode}
We again use \coq{RecF_inv} to obtain the structurally smaller proof to use in
the recursive calls. As we did with catamorphisms and recursive anamorphisms, we
prove that \coq{hylo_def} is respectful, and use this proof to built the
corresponding higher-order proper morphism:
\begin{coqcode}
Alg F B ~> RCoalg F A ~> A ~> B
\end{coqcode}
Finally, we show that recursive hylomorphisms are the unique solution to the
hylomorphism equation.
\begin{coqcode}
Lemma hylo_univ (g : Alg F B) (h : RCoalg F A) (f : A ~> B)
    : f =e hylo g h <-> f =e g \o fmap f \o h.
\end{coqcode}
Hylomorphisms fusion falls out from the universal property:
\begin{coqcode}
Lemma hylo_fusion_l (h1 : RCoalg F A) (g1 : Alg F B) (g2 : Alg F C)
  (f2 : B ~> C) (E2 : f2 \o g1 =e g2 \o fmap f2) : f2 \o hylo g1 h1 =e hylo g2 h1.
\end{coqcode}
% Qed.
%   rewrite hylo_univ, fmap_comp, compA, <- E2,
%           <- compA, <- compA, (compA g1), <- hylo_unr.  reflexivity.

  % rewrite hylo_univ, fmap_comp, compA, <- E2,
  %         <- compA, <- compA, (compA g1), <- hylo_unr.  reflexivity.

  % (* f2 \o hylo g1 h1 =e hylo g2 h1 *)
  % (* f2 \o hylo g1 h1 =e (g2 \o fmap (f2 \o hylo g1 h1)) \o h1 *)
  % (* f2 \o hylo g1 h1 =e (g2 \o (fmap f2 \o fmap (hylo g1 h1))) \o h1 *)
  % (* f2 \o hylo g1 h1 =e ((g2 \o fmap f2) \o fmap (hylo g1 h1)) \o h1 *)
  % (* f2 \o hylo g1 h1 =e ((f2 \o g1) \o fmap (hylo g1 h1)) \o h1 *)
  % (* f2 \o hylo g1 h1 =e (f2 \o g1) \o (fmap (hylo g1 h1) \o h1) *)
  % (* f2 \o hylo g1 h1 =e f2 \o (g1 \o (fmap (hylo g1 h1) \o h1)) *)
  % (* f2 \o hylo g1 h1 =e f2 \o ((g1 \o (fmap (hylo g1 h1)) \o h1) *)
  % (* f2 \o hylo g1 h1 =e f2 \o hylo g1 h1 *)
Using the hylo fusion law, we can prove the well-known \emph{deforestation
optimisation}. This is when two consecutive recursive computations, one that
builds a data structure, and another one that consumes it, can be fused together
into a single recursive definition. This, in turn, allows us to prove that a
recursive hylomorphism is the composition of a catamorphism and a recursive
anamorphism.
\begin{coqcode}
Lemma deforest (h1 : RCoalg F A) (g2 : Alg F C)
  (g1 : Alg F B) (h2 : RCoalg F B) (INV: h2 \o g1 =e id)
  : hylo g2 h2 \o hylo g1 h1 =e hylo g2 h1.
\end{coqcode}

\subsection{Recursive Anamorphisms}
Anamorphisms specialise to an hylomorphism into an inductive data type.
Intuitively, this is because a recursive coalgebra can be applied only finitely
many times, so when this is applied to an anamorphism the only possible traces
it can produce are the finite ones.  We show then that an anamorphisms applied
to a recursive coalgebra is in one-to-one correspondence with an hylomorphism
applied to the same recursive coalgebra and the \coq{In} map which is a map into
the inductive data type defined by \coq{In}.

To do this we first define a predicate which holds if and only if the element of
a coinductive data type is finite:
\begin{coqcode}
Inductive FinF : GFix F -> Prop :=
| FinF_fold (x : GFix F) : (forall e, FinF (cont (g_out x) e)) -> FinF x.
\end{coqcode}
Catamorphisms from this initial algebra are denoted by \coq{ccata}, and they can
be used to define the isomorphism between \coq{{x : GFix F | FinF x}}
and \coq{LFix F}. In particular:
\begin{coqcode}
Lemma cata_ccata `{setoid A} (f : Alg F A) : cata f \o ccata l_in =e ccata f.
Lemma ccata_cata `{setoid A} (f : Alg F A) : ccata f \o cata lg_in =e cata f.
\end{coqcode}
Using this new initial algebra, we can compose catamorphisms and anamorphisms,
\emph{as long as the anamorphisms produce finite trees}. We refer the reader to
the supplementary material and Coq code to see how we encode this. In this
paper, however, we focus on the fact that, if anamorphisms produce finite trees,
then their return type can also be the \emph{least} fixed point of the container
extension. We encode in Coq the notion of \emph{recursive anamorphisms} and
\emph{recursive coalgebras}~\cite{AdamekMM19,CaprettaUV04}.

\begin{coqcode}
Corollary cata_ana_hylo `(F : Cont Sh P) `{setoid A} `{setoid B}
  (g : Alg F B) (h : RCoalg F A)
: cata g \o rana h =e hylo g h.
\end{coqcode}
This result will allow us to \emph{calculate} programs by providing small, easy
to verify implementations, and combining them into efficient recursive functions
by \emph{using Coq's proofs of hylomorphism laws as optimisation passes}.

This predicate is equivalent to the proposition that the anamorphism of such
coalgebras produce only finite trees:
\begin{coqcode}
Lemma ana_rec_term `{setoid A} (h : Coalg F A)
  : (forall x, FinF (ana h x)) <-> (forall x, RecF h x).
\end{coqcode}

\mpav{Moved here}

Note that such proofs will be erased when extracting code to OCaml, and this is
standard practice when building general recursive functions in
Coq\footnote{\dcas{This technique had a name in the community, but I cannot
remember. Something predicates}}. The core definitions are the following:
\begin{coqcode}
Lemma RecF_inv (h : Coalg F A) x : RecF h x -> forall e, RecF h (cont (h x) e).
